Simulating High-Dimensional Multivariate Data using the Bigsimr R Package

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr.Rmd
repo arXiv docs

---
title: Simulating High-Dimensional Multivariate Data using the Bigsimr R Package
authors:
  - name: A. Grant Schissler
    department: Department of Mathematics \& Statistics
    affiliation: University of Nevada, Reno
    location: Reno, NV 89557
    email: grant.schissler@gmail.com
  - name: Edward J. Bedrick
    affiliation: University of Arizona
  - name: Alexander D. Knudson
    affiliation: University of Nevada, Reno \AND
  - name: Tomasz J. Kozubowski
    affiliation: University of Nevada, Reno
  - name: Tin Nguyen
    affiliation: University of Nevada, Reno
  - name: Anna K. Panorska
    affiliation: University of Nevada, Reno \AND
  - name: Juli Petereit
    affiliation: University of Nevada, Reno
  - name: Walter W. Piegorsch
    affiliation: University of Arizona
  - name: Duc Tran
    affiliation: University of Nevada, Reno
abstract: |
  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments. Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To close this gap, we developed the Bigsimr Julia package with R and Python interfaces. This paper focuses on the R interface. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. bigsimr contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
keywords:
  - high-dimensional data
  - multivariate simulation
  - nonparametric correlation
  - Gaussian copula
  - RNA-sequencing data
  - breast cancer
bibliography: references.bib
biblio-style: unsrt
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

>>>>

---
title: Simulating High-Dimensional Multivariate Data using the Bigsimr R Package
authors:
  - name: A. Grant Schissler
    department: Department of Mathematics \& Statistics
    affiliation: University of Nevada, Reno
  - name: Edward J. Bedrick
    department: Epidemiology and Biostatistics Department
    affiliation: University of Arizona
  - name: Alexander D. Knudson
    department: Department of Mathematics \& Statistics
    affiliation: University of Nevada, Reno
  - name: Tomasz J. Kozubowski
    department: Department of Mathematics \& Statistics
    affiliation: University of Nevada, Reno
  - name: Tin Nguyen
    department: Department of Computer Science \& Engineering
    affiliation: University of Nevada, Reno
  - name: Anna K. Panorska
    department: Department of Mathematics \& Statistics
    affiliation: University of Nevada, Reno
  - name: Juli Petereit
    department: Nevada Bioinformatics Center
    affiliation: University of Nevada, Reno
  - name: Walter W. Piegorsch
    department: Mathematics Department
    affiliation: University of Arizona
  - name: Duc Tran
    department: Department of Computer Science \& Engineering
    affiliation: University of Nevada, Reno
abstract: |
  It is critical to accurately simulate data when employing Monte Carlo techniques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained in high-throughput biomedical experiments. Due to the computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. To close this gap, we developed the Bigsimr Julia package with R and Python interfaces. This paper focuses on the R interface. These packages empower high-dimensional random vector simulation with arbitrary marginal distributions and dependency via a Pearson, Spearman, or Kendall correlation matrix. bigsimr contains high-performance features, including multi-core and graphical-processing-unit-accelerated algorithms to estimate correlation and compute the nearest correlation matrix. Monte Carlo studies quantify the accuracy and scalability of our approach, up to $d=10,000$. We describe example workflows and apply to a high-dimensional data set --- RNA-sequencing data obtained from breast cancer tumor samples.
keywords:
  - high-dimensional data
  - multivariate simulation
  - nonparametric correlation
  - Gaussian copula
  - RNA-sequencing data
  - breast cancer
bibliography: references.bib
biblio-style: unsrt
output:
  rticles::arxiv_article:
    includes:
      in_header: preamble.tex
---

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr_submission.tex
repo arXiv docs

\author{
    A. Grant Schissler
   \\
    Department of Mathematics \& Statistics \\
    University of Nevada, Reno \\
  Reno, NV 89557 \\
  \texttt{\href{mailto:grant.schissler@gmail.com}{\nolinkurl{grant.schissler@gmail.com}}} \\
   \And
    Edward J. Bedrick
   \\
    University of Arizona \\
   \\
  \texttt{} \\
   \And
    Alexander D. Knudson
   \\
    University of Nevada, Reno \AND \\
   \\
  \texttt{} \\
   \And
    Tomasz J. Kozubowski
   \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Tin Nguyen
   \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Anna K. Panorska
   \\
    University of Nevada, Reno \AND \\
   \\
  \texttt{} \\
   \And
    Juli Petereit
   \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Walter W. Piegorsch
   \\
    University of Arizona \\
   \\
  \texttt{} \\
   \And
    Duc Tran
   \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
  }

@@

@@

\begin{verbatim}
## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
## (lower = -0.8807315594247829, upper = 0.880286004475508)
\end{verbatim}

>>>>

\author{
    A. Grant Schissler
   \\
    Department of Mathematics \& Statistics \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Edward J. Bedrick
   \\
    Epidemiology and Biostatistics Department \\
    University of Arizona \\
   \\
  \texttt{} \\
   \And
    Alexander D. Knudson
   \\
    Department of Mathematics \& Statistics \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Tomasz J. Kozubowski
   \\
    Department of Mathematics \& Statistics \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Tin Nguyen
   \\
    Department of Computer Science \& Engineering \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Anna K. Panorska
   \\
    Department of Mathematics \& Statistics \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Juli Petereit
   \\
    Nevada Bioinformatics Center \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
   \And
    Walter W. Piegorsch
   \\
    Mathematics Department \\
    University of Arizona \\
   \\
  \texttt{} \\
   \And
    Duc Tran
   \\
    Department of Computer Science \& Engineering \\
    University of Nevada, Reno \\
   \\
  \texttt{} \\
  }

@@

© A.G. Schissler. This is an Open Access article distributed under the terms of the Creative Commons License Attribution-NonCommercial-ShareAlike 4.0 International (CC BY-NC-SA 4.0) (https://creativecommons.org/licenses/by-nc-sa/4.0/), which permits reusers to distribute, remix, adapt, and build upon the material in any medium or format for noncommercial purposes only, and only so long as attribution is given to the creator. If you remix, adapt, or build upon the material, you must license the modified material under identical terms.

@@

\begin{verbatim}
## Julia Object of type NamedTuple{(:lower, :upper),Tuple{Float64,Float64}}.
## (lower = -0.8829854580269612, upper = 0.882607249144551)
\end{verbatim}

### 2022-01-07 arXiv_bigsimr/arXiv_bigsimr_submission/arXiv_bigsimr_submission.tex
repo arXiv docs

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch010-realDataFig-1.pdf}
\caption{\label{fig:ch010-realDataFig}Pairwise scatterplots for three
example genes (Table 1), with estimated Spearman's correlations and
marginal density estimates displayed. The data possess outliers,
heavy-right tails, are discrete, and have non-trivial intergene
correlations. Modeling these data motivate our simulation methodology.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-aq-joint-dist-1.pdf}
\caption{\label{fig:ch030-aq-joint-dist}Bivariate scatterplot of Ozone
versus Temp with estimated marginal densities. The Ozone data are
modeled marginally as log-normal and the Temperature data as normal.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch030-plot-sim-1.pdf}
\caption{\label{fig:ch030-plot-sim}Contour plot and marginal densities
for the simulated bivariate distribution of Air Quality Temperatures and
Ozone levels. The simulated points mimic the observed data with respect
to both the marginal characteristics and bivariate association.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNormPlot-1.pdf}
\caption{\label{fig:ch040-biNormPlot}\texttt{bigsimr} recovers the
Pearson specified correlations for MVN.}
\end{figure}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biGammaPlot-1.pdf}
\caption{\label{fig:ch040-biGammaPlot}\texttt{bigsimr} recovers the
Pearson specified correlations for Bivariate Gamma.}
\end{figure}

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-biNegBinPlot-1.pdf}
\caption{\label{fig:ch040-biNegBinPlot}\texttt{bigsimr} recovers the
correlations for bivariate negative binomial only approximately for
Pearson but (nearly) exactly for the rank-based correlations.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-bPlot-1.pdf}
\caption{\label{fig:ch040-bPlot}Bivariate simulations match target
correlations across the entire range of feasible correlations. The
horizontal axis plots the specified target correlations for each
bivariate margin. Normal margins are plotted in dark dark grey, gamma in
medium grey, and negative binomial in light grey. As the number of
simulated vectors \(B\) increases from left to right, the variation in
estimated correlations (vertical axis) decreases. The dashed line
indicates equality between the specified and estimated correlations.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch040-largeDfig-1.pdf}
\caption{\label{fig:ch040-largeDfig}Computation times for HD
multivariate gamma simulation.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{arXiv_bigsimr_files/figure-latex/ch050-simDataFig-1.pdf}
\caption{\label{fig:ch050-simDataFig}Simulated data for three selected
high-expressing genes generally replicates the estimated data structure.
The data do not exhibit outlying points, but do possess the desired
Spearman correlations, central tendencies, and discrete values.}
\end{figure}

\begin{figure}
\includegraphics[width=0.8\linewidth]{../fig/ch050-figBRCA} \caption{\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
\end{figure}

>>>>

\begin{figure}
\centering
\includegraphics{ch010-realDataFig-1.pdf}
\caption{\label{fig:ch010-realDataFig}Pairwise scatterplots for three
example genes (Table 1), with estimated Spearman's correlations and
marginal density estimates displayed. The data possess outliers,
heavy-right tails, are discrete, and have non-trivial intergene
correlations. Modeling these data motivate our simulation methodology.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch030-aq-joint-dist-1.pdf}
\caption{\label{fig:ch030-aq-joint-dist}Bivariate scatterplot of Ozone
versus Temp with estimated marginal densities. The Ozone data are
modeled marginally as log-normal and the Temperature data as normal.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch030-plot-sim-1.pdf}
\caption{\label{fig:ch030-plot-sim}Contour plot and marginal densities
for the simulated bivariate distribution of Air Quality Temperatures and
Ozone levels. The simulated points mimic the observed data with respect
to both the marginal characteristics and bivariate association.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch040-biNormPlot-1.pdf}
\caption{\label{fig:ch040-biNormPlot}\texttt{bigsimr} recovers the
Pearson specified correlations for MVN.}
\end{figure}

\begin{figure}
\centering
\includegraphics{ch040-biGammaPlot-1.pdf}
\caption{\label{fig:ch040-biGammaPlot}\texttt{bigsimr} recovers the
Pearson specified correlations for Bivariate Gamma.}
\end{figure}

\begin{figure}
\centering
\includegraphics{ch040-biNegBinPlot-1.pdf}
\caption{\label{fig:ch040-biNegBinPlot}\texttt{bigsimr} recovers the
correlations for bivariate negative binomial only approximately for
Pearson but (nearly) exactly for the rank-based correlations.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch040-bPlot-1.pdf}
\caption{\label{fig:ch040-bPlot}Bivariate simulations match target
correlations across the entire range of feasible correlations. The
horizontal axis plots the specified target correlations for each
bivariate margin. Normal margins are plotted in dark dark grey, gamma in
medium grey, and negative binomial in light grey. As the number of
simulated vectors \(B\) increases from left to right, the variation in
estimated correlations (vertical axis) decreases. The dashed line
indicates equality between the specified and estimated correlations.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch040-largeDfig-1.pdf}
\caption{\label{fig:ch040-largeDfig}Computation times for HD
multivariate gamma simulation.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{ch050-simDataFig-1.pdf}
\caption{\label{fig:ch050-simDataFig}Simulated data for three selected
high-expressing genes generally replicates the estimated data structure.
The data do not exhibit outlying points, but do possess the desired
Spearman correlations, central tendencies, and discrete values.}
\end{figure}

\begin{figure}
\includegraphics[width=0.8\linewidth]{ch050-figBRCA} \caption{\label{fig:ch050-figBRCA}Simulated random vectors from a multivariate negative binomial replicate the estimated structure from an RNA-seq data set. The dashed red lines indicate equality between estimated parameters from simulated data (vertical axes) and the specified target parameters (horizontal axes).}\label{fig:ch050-figBRCA}
\end{figure}
